---
title: "Hospital Quality Project"
output: md_document

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
In this project I work with dataset from The [Hosptial Compare web site](http://hospitalcompare.hhs.gov/) run by the U.S. Department of Health and Human Services.
The web site provide data and
information about the quality of care at over 4,000 Medicare-certied hospitals in the U.S.

For this project we use only a subset of data which the web site contains.

There are three files:

- **outcome-of-care-measures.csv**: Contains information about 30-day mortality and readmission rates
for heart attacks, heart failure, and pneumonia for over 4,000 hospitals.
- **hospital-data.csv**: Contains information about each hospital.
- **Hospital_Revised_Flatfiles.pdf**: Descriptions of the variables in each file (i.e the code book).

### what we are going to do in this assignment:

1- Plot the 30-day mortality rates for heart attack

2- Finding the best hospital in a state

3- Ranking hospitals by outcome in a state

4- Ranking hospitals in all states

-------------------------------------------------------------------------------------------------
-------------------------------------------------------------------------------------------------

## 1-Histogram for the 30-day mortality rates for heart attack
In this step I read the data from outcome-of-care-measures.csv ,then plot a histogram for column(11) which is contain the data for the 30-day mortality rates for heart attack.

```{r}
data <- read.csv("outcome-of-care-measures.csv", colClasses = "character")
data[,11] <- as.numeric(data[, 11])
outcome_data <- data[, 11]
hist(outcome_data, main = "The 30-day mortality rates for heart attack",
     xlab="Death rates from heart attack",
     col = "darkmagenta")
```


------------------------------------------------------------------------------------------------
------------------------------------------------------------------------------------------------


## 2-Function returns the best hospital in a state
A function called "best" that take two arguments: the 2-character abbreviated name of a state and an outcome name.
The function reads the outcome-of-care-measures.csv file and returns a character vector
with the name of the hospital that has the best (lowest) 30-day mortality for the specified outcome in that state.

**Note:**
The outcomes can
be one of "heart attack", "heart failure", or "pneumonia".

### Implementation:
```{r}
best <- function(state, outcome)
{
  data <- read.csv("outcome-of-care-measures.csv", colClasses = "character")
  
  #First check outcome if is valid
  expected_outcome <- c("heart attack", "heart failure", "pneumonia")
  check_outcome <- outcome %in% expected_outcome
  if (check_outcome == FALSE )
  {
    stop("invalid outcome")
  }
  #check state if is valid 
  check_state <- state %in% data$State
  if (check_state == FALSE )
  {
    stop("invalid state")
  }
  
  
  #change columns names
  data <- data[c(2, 7, 11, 17, 23)]
  names(data)[1] <- "name"
  names(data)[3] <- "heart attack"
  names(data)[4] <- "heart failure"
  names(data)[5] <- "pneumonia"
  
  data <- data[data$State == state & data[outcome] != "Not Available",]
  outcome_data <- data[, outcome]  #Read the outcome data
  
  
  
  #Return the best hospital
  index <- which.min(outcome_data)  #return index of the minmum value
  return(data[index, ]$name)
}
```

**Test cases:**
```{r eval=best , error=TRUE}
best("TX", "heart attack")
best("TX", "heart failure")
best("BB", "heart attack")
best("NY", "hert attack")


```


------------------------------------------------------------------------------------------------
------------------------------------------------------------------------------------------------

## 3-Function ranking hospitals by outcome in a state
A function called rankhospital that takes three arguments: the 2-character abbreviated name of a
state (state), an outcome (outcome), and the ranking of a hospital in that state for that outcome (num).

The function returns a character vector with the name
of the hospital that has the ranking specified by the num argument.

### Implementation:
```{r}
rankhospital <- function(state, outcome, num = "best")
{
  
  data <- read.csv("outcome-of-care-measures.csv", colClasses = "character")
  
  #First check outcome if is valid
  expected_outcome <- c("heart attack", "heart failure", "pneumonia")
  check_outcome <- outcome %in% expected_outcome
  if (check_outcome == FALSE )
  {
    stop("invalid outcome")
  }
  #check state if is valid 
  check_state <- state %in% data$State
  if (check_state == FALSE )
  {
    stop("invalid state")
  }
  
  
  #change columns names
  data <- data[c(2, 7, 11, 17, 23)]
  names(data)[1] <- "name"
  names(data)[3] <- "heart attack"
  names(data)[4] <- "heart failure"
  names(data)[5] <- "pneumonia"
  
  data <- data[data$State == state & data[outcome] != "Not Available",]
  
  #Ranking
  data[outcome] <- as.data.frame(sapply(data[outcome], as.numeric))
  data <- data[order(data$name, data[outcome] ,decreasing = FALSE), ]

  outcome_data <- data[, outcome]  #Read the outcome data
  
  if( num == "best" ) {
    index <- which.min(outcome_data)
  } else if( num == "worst" ) {
    index <- which.max(outcome_data)
  } else {
    index <- num
  }
  
  return(data[index, ]$name)
  
}
```

**Test cases:**
```{r eval=rankhospital , error=TRUE}
rankhospital("TX", "heart failure", 4)
rankhospital("MD", "heart attack", "worst")
rankhospital("MN", "heart attack", 5000)


```


------------------------------------------------------------------------------------------------
------------------------------------------------------------------------------------------------

## 4-Function ranking hospitals in all states
A function called rankall that takes two arguments: an outcome name (outcome) and a hospital ranking (num), returns a 2-column data frame
containing the hospital in each state that has the ranking specified in num.

### Implementation

```{r}
rankall <- function(outcome, num = "best")
{
  data <- read.csv("outcome-of-care-measures.csv", colClasses = "character")
  
  #First check outcome if is valid
  expected_outcome <- c("heart attack", "heart failure", "pneumonia")
  check_outcome <- outcome %in% expected_outcome
  if (check_outcome == FALSE )
  {
    stop("invalid outcome")
  }
  
  #change columns names
  data <- data[c(2, 7, 11, 17, 23)]
  names(data)[1] <- "name"
  names(data)[2] <- "state"
  names(data)[3] <- "heart attack"
  names(data)[4] <- "heart failure"
  names(data)[5] <- "pneumonia"
  
  data <- data[data[outcome] != "Not Available",]

  data[outcome] <- as.data.frame(sapply(data[outcome], as.numeric))
  data <- data[order(data$name, data[outcome] ,decreasing = FALSE), ]
  
  #Helper Function
  hospital_rank <- function(df, s, n) 
  {
    df <- df[df$state==s, ]
    outcome_data <- df[,outcome]
    if( n == "best" ) {
      index <- which.min(outcome_data)
    }else if( n == "worst" ) {
      index <- which.max(outcome_data)
    }else {
      index <- n
    }
    df[index, ]$name
  }
  
  ## For each state, find the hospital of the given rank
  states <- data[, 2]
  states <- unique(states)
  ranking_data <- data.frame(hospital = c(), state = c())
  for(st in states) {
    hosp <- hospital_rank(data, st, num)
    ranking_data <- rbind(ranking_data, data.frame(hospital = hosp, state = st))
  }

  ## Return a data frame(arranged) with hospital name and state name
  ranking_data <- ranking_data[order(ranking_data['state'], decreasing = FALSE), ]
  return(ranking_data)
  
}
```

**Test cases:**
```{r eval=rankall, error=TRUE}
head(rankall("heart attack", 20), 10)
tail(rankall("pneumonia", "worst"), 3)
tail(rankall("heart failure"), 10)

```










