---
title: Explore the NOAA Storm Database to determine the effects of weather events
  on personal and property in US.
output: html_document
  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
--------------------------------------------------------------------------
--------------------------------------------------------------------------
## 1.Synopsis

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.


This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

------------------------------------------------------------------------------------------------------------

## 2.Data Processing

### 2.1 Data
- The data for this assignment come in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size.

- You can download the [Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2).

- If you need more information about the Storm database:
    - National Weather Service [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf).
    
    - National Climatic Data Center Storm Events [FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf).
    
- The events in the database start in the year 1950 and end in November 2011.

- In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records,More recent years should be considered more complete.

### 2.2 Data Processing
#### 2.2.1 Loading the Data

Loading neccessarry packages.

```{r}
library(dplyr)
library(ggplot2)
library(gridExtra)
```

Loading DataSet

```{r load_data, cache= TRUE}
before <- Sys.time()  # To determine the time before reading data
storm_data <- read.csv("repdata_data_StormData.csv")
after <- Sys.time()  #  determine the time after reading data

loading_time <- after - before

print(paste("The time was taking to reading data is ", loading_time))
 
```

#### 2.2.2 
#### Data processing

```{r}
head(storm_data)

```

In this analysis, we only focus on analyzing the effects of weather events to public health and economy,
So we Select columns related to these.
```{r}
storm_data <- storm_data[, c("BGN_DATE", "EVTYPE", "FATALITIES", "INJURIES", 
    "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")]
```

### First, According United States We want to to determine which types of events are most harmful with respect to population health?

**The following lines of code summarize the injury and fatality data per event.**

```{r}
event_fatality <- storm_data %>% group_by(EVTYPE) %>% summarize(sum(FATALITIES,na.rm=TRUE))
colnames(event_fatality) <- c("EVTYPE","FATALITIES")
event_injuries <- storm_data %>% group_by(EVTYPE) %>% summarize(sum(INJURIES,na.rm=TRUE))
colnames(event_injuries) <- c("EVTYPE","INJURIES")
```

**Then rearrange the data in descending order to determine the most harmful events.**

```{r}
event_fatality <- event_fatality[order(event_fatality$FATALITIES,decreasing = TRUE),]
event_injuries <- event_injuries[order(event_injuries$INJURIES,decreasing = TRUE),]
```

#### Results

**Getting and Plotting the top 7 causes of fatality and injury.**

```{r}
event_fatality <- event_fatality[1:7,]
event_injuries <- event_injuries[1:7,]
injPlot <- ggplot(event_injuries) + geom_bar(aes(x = reorder(EVTYPE, -INJURIES), y = INJURIES,        
    fill=interaction(INJURIES,EVTYPE)), show.legend=FALSE, stat="identity") + 
    xlab("Event")+ylab("Injuries")+ggtitle("Top 7 Events Causing Injury") + scale_fill_brewer(palette="Spectral")

fatPlot <- ggplot(event_fatality)+ geom_bar(aes(x = reorder(EVTYPE, -FATALITIES), y = FATALITIES, 
    fill=interaction(FATALITIES,EVTYPE)),show.legend=FALSE,stat="identity") + 
    xlab("Event")+ylab("Fatalities")+ggtitle("Top 7 Events Causing Fatalities") +     
    scale_fill_brewer(palette="Spectral")
grid.arrange(fatPlot,injPlot)
```

**NOTE:From the plot we noticed the "Tornados" are the leading cause  of injuries and death.**

--------------------------------------------------------------------------------------------------------

### Second, We want to determine which types of events have the greatest economic consequences?
#### Data Processing
#### Converting economic data exponents to values then multiplying it to actual values.

```{r}
storm_data$PROPDMGEXP <- gsub("m",1000000,storm_data$PROPDMGEXP,ignore.case = TRUE)
storm_data$PROPDMGEXP <- gsub("k",1000,storm_data$PROPDMGEXP,ignore.case = TRUE)
storm_data$PROPDMGEXP <- gsub("b",1000000000,storm_data$PROPDMGEXP,ignore.case = TRUE)
storm_data$CROPDMGEXP <- gsub("m",1000000,storm_data$CROPDMGEXP,ignore.case=TRUE)
storm_data$CROPDMGEXP <- gsub("k",1000,storm_data$CROPDMGEXP,ignore.case=TRUE)
storm_data$CROPDMGEXP <- gsub("b",1000000000,storm_data$CROPDMGEXP,ignore.case=TRUE)
storm_data_complete_case <- storm_data[complete.cases(storm_data),]
storm_data_complete_case$CROPDMGEXP <- as.numeric(storm_data_complete_case$CROPDMGEXP)


storm_data_complete_case$PROPDMGEXP <- as.numeric(storm_data_complete_case$PROPDMGEXP)
storm_data_complete_case$TOTALPROPDMG <- storm_data_complete_case$PROPDMG * storm_data_complete_case$PROPDMGEXP
storm_data_complete_case$TOTALCROPDMG <- storm_data_complete_case$CROPDMG * storm_data_complete_case$CROPDMGEXP
```

**Summarizing and re-organizing data to prepare it for Plotting.**

```{r}
storm_data_complete_case_prop <- storm_data_complete_case %>% group_by(EVTYPE) %>% summarize(grandtotalprop=sum(TOTALPROPDMG,na.rm=TRUE))
storm_data_complete_case_crop <- storm_data_complete_case %>% group_by(EVTYPE) %>% summarize(grandtotalcrop=sum(TOTALCROPDMG,na.rm=TRUE))
total_eco_dmg <- full_join(storm_data_complete_case_crop,storm_data_complete_case_prop,by="EVTYPE")
total_eco_dmg$Total <- total_eco_dmg$grandtotalcrop+total_eco_dmg$grandtotalprop
total_crop_dmg <- total_eco_dmg[order(total_eco_dmg$grandtotalcrop,decreasing=TRUE),]
total_crop_dmg <- total_crop_dmg[1:7,]
total_prop_dmg <- total_eco_dmg[order(total_eco_dmg$grandtotalprop,decreasing = TRUE),]
total_prop_dmg <- total_prop_dmg[1:7,]
total_eco_dmg <- total_eco_dmg[order(total_eco_dmg$Total,decreasing = TRUE),]
total_eco_dmg <- total_eco_dmg[1:7,]
```

### Results

**Plotting and seeing top events causing the most damage economically**

```{r}
cropPlot <- ggplot()+geom_bar(data=total_crop_dmg,aes(x = reorder(EVTYPE, -grandtotalcrop),y=grandtotalcrop,fill=interaction(EVTYPE,grandtotalcrop)),show.legend = FALSE,stat="identity")+
    xlab("Event")+ylab("Economic Damage")+ggtitle("Top 7 Events causing Damage (Crop)")+
    scale_fill_brewer(palette="BrBG")


propPlot <- ggplot()+geom_bar(data=total_prop_dmg,aes(x = reorder(EVTYPE, -grandtotalprop),y=grandtotalprop,fill=interaction(EVTYPE,grandtotalprop)),show.legend = FALSE,stat="identity")+
    xlab("Event")+ylab("Economic Damage")+ggtitle("Top 7 Events causing Damage (Property)") +
    scale_fill_brewer(palette="BrBG")

totalPlot <- ggplot()+geom_bar(data=total_eco_dmg,aes(x = reorder(EVTYPE,-Total),y = Total,fill=interaction(EVTYPE,Total)),show.legend = FALSE,stat="identity")+
    xlab("Event")+ylab("Economic Damage")+ggtitle("Top 7 Events causing Damage (Total)")+
    scale_fill_brewer(palette="BrBG")

grid.arrange(cropPlot,propPlot,totalPlot)

```

### 3.Conclusion
Tornados caused the maximum number of fatalities and injuries. It was followed by Excessive Heat for fatalities and Thunderstorm wind for injuries.

Floods caused the maximum property damage where as Drought caused the maximum crop damage. Second major events that caused the maximum damage was Hurricanes/Typhoos for property damage and Floods for crop damage.








